---
title: "More Examples"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: SCI 2000--Introduction to Data Science
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(cache = FALSE, message = FALSE,
                      linewidth = 50)
```

## Explore data {.allowframebreaks}

  - We will use data on Forced Expiratory Volume (FEV) in children age 3 to 19 from East Boston recorded during the 1970s.
    + Can be downloaded from https://hbiostat.org/data/, but I also added a copy on UM Learn.
  - The dataset contains information on age, height, sex, and smoking status.
  - **Outcome**: FEV
  
```{r, message = FALSE}
library(tidyverse)
# Import dataset into R
data_fev <- read_csv("FEV.csv")
glimpse(data_fev, width = 50) # So it fits the slide

# Explore data
ggplot(data_fev, aes(x = sex, y = fev)) +
  geom_boxplot()
ggplot(data_fev, aes(x = smoke, y = fev)) +
  geom_boxplot()
```

```{r}
# Smokers have higher FEV??
fit <- lm(fev ~ smoke, data = data_fev)
coef(fit)
confint(fit)
```

  - Non-smokers have, on average, an FEV measure that is 0.7 lower than smokers... What can be going on?

```{r, message = FALSE}
# Look at FEV vs age and height
ggplot(data_fev, aes(x = age, y = fev)) +
  geom_point()
ggplot(data_fev, aes(x = height, y = fev)) +
  geom_point()
```

  - The association between FEV and smoking status is **spurious**: it looks like it is driven by the fact that:
    - Older children are taller, have larger lungs, and therefore higher FEV.
    - Older children are more likely to be smokers.
  - We also say that age and height are **confounders** for the association between FEV and smoking status.
    
## Fit a linear model {.allowframebreaks}

  - The idea is that we are comparing older and younger children together, thus creating this spurious association.
    + What if we only compared children of the same age?
  - Linear regression actually allows us to **adjust** for the effect of age and height on FEV.

```{r}
# Fit linear model
model <- lm(fev ~ smoke + sex + age + height, 
            data = data_fev)
```

\vspace{1cm}

```{r}
coef(model)
```

\vspace{2cm}

```{r}
confint(model)
```

  - Non-smokers have, on average, an FEV measure that is 0.08 *higher* than smokers, when adjusting for age, height and sex.
    + And it's no longer significant (0 is in the confidence interval)
  
## Residual analysis {.allowframebreaks}
  
```{r}
library(broom)
# Plot outcome vs fitted values
augment(model) %>% 
  ggplot(aes(x = .fitted, y = fev)) +
  geom_point() +
  geom_abline(intercept = 0,
              slope = 1)
```

```{r eval = FALSE}
# Can also colour points according to smoking status
augment(model) %>% 
  ggplot(aes(x = .fitted, y = fev)) +
  geom_point(aes(colour = smoke)) +
  geom_abline(intercept = 0,
              slope = 1)
```

```{r}
# Plot residuals vs fitted values
augment(model) %>% 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0,
             linetype = "dotted")
```

  - We found evidence that additivity/linearity is not met.
    + Outcome vs fitted plot.
    + Given our data visualizations, it is likely that relationship between FEV and height is nonlinear.
  - We found evidence of unequal variance.
    + Residual vs fitted values: higher variance with larger fitted values.
  - How can we use residual analysis to decide how we could improve the model?

## Fit a second linear model {.allowframebreaks}

```{r}
# We will add a quadratic term for height
# The function I() protects height^2
# Try removing it from the code below and 
# see how it's different
model2 <- lm(fev ~ smoke + sex + age + 
               height + I(height^2), 
            data = data_fev)
```

\vspace{1cm}

```{r}
coef(model2)
```

\vspace{2cm}

```{r linewidth = 60}
confint(model2)
```

  - Non-smokers have, on average, an FEV measure that is 0.13 *higher* than smokers, when adjusting for age, height and sex.
    + And now it's back to being significant

## Residual analysis redux {.allowframebreaks}
  
```{r}
augment(model2) %>% 
  ggplot(aes(x = .fitted, y = fev)) +
  geom_point() +
  geom_abline(intercept = 0,
              slope = 1)
```

```{r}
augment(model2) %>% 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0,
             linetype = "dotted")
```

<center>
We still have evidence of unequal variance, but at least additivity/linearity now seem to hold!
</center>

```{r linewidth = 60}
# Let's use robust standard errors
library(lmtest)
library(sandwich)
coefci(model2, vcov. = vcovHC(model2))
```

## Summary 

  - We still have the same interpretation for our regression coefficient:
    + Non-smokers have, on average, an FEV measure that is 0.13 *higher* than smokers, when adjusting for age, height and sex.
  - With the robust standard errors, the confidence interval is wider, and so the association between FEV and smoking status (accounting for age, height and sex) is no longer significant.
  - Because we are now confident our assumptions hold, the right conclusion from our analysis is the one based on the last model.
    + Quadratic term for height
    + Robust standard errors
