---
title: "Introduction to Relational Databases"
draft: true
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: SCI 2000--Introduction to Data Science
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(cache = FALSE, message = FALSE,
                      linewidth = 50)
```

## Lecture Objectives

  - Learn the main features of a relational database.
  - Query data from such a database using `R`.'
  - Explain the differences between importing data in `R` and querying a relational database.

## Motivation

  - Last week, we talked about relational data.
    + Related datasets with a variable in common encoding the relationship.
  - Most businesses collect relational data.
    + Employee data and scheduling information
    + Product, client and order data
  - **Relational databases** are the main way of efficiently storing this relational data.
    + For easy maintenance and accuracy.
    + For fast retrieval.
    
## Disclaimer

  - There is no way we can have a thorough discussion of relational databases in SCI 2000. 
    + If you're interested, you should look at COMP 3380
  - Topics we will **not** discuss:
    + SQL (Structured Query Language)
    + Database design
    + Relational algebra
    
## Basic vocabulary

  - A **relational database** is a collection of tables.
  - A **table** is a collection of records.
  - A **record** is a collection of attributes.
  - An **attribute** is a piece of information that we want to capture.
    + Could be a string, a date-time, an integer, a floating point number, etc 
    + Could even be an image or an audio file.
  - Typically, each record has a special attribute called a **primary key** to uniquely identify it within the table, and to refer to it in a *different* table.
    + If a record contains the primary key of another table, it is called a **foreign key**.
    
## Example

| Product code | Description       | Price |
|--------------|-------------------|-------|
| A416         | Nails, box        | 0.14$ |
| C923         | Drawing pins, box | 0.08$ |

  - This is a table, with two records. 
  - Each record has three attributes, and "Product code" is the primary key.
  
----

| Invoice code | Invoice line | Product code | Quantity |
|--------------|--------------|--------------|----------|
| 3804         | 1            | A416         | 10       |
| 3804         | 2            | C923         | 15       |

  - This is another table, also with two records. 
  - Each record has four attributes, and "Product code" is a foreign key.
  - This table does **not** have a primary key (because "Invoice code" is not unique).
  
----

  - You could imagine a separate table containing information specific to each invoice.
    + E.g. Date of the order, customer number, invoice total, whether it has been paid, etc.
  - Similarly, you could imagine another table containing information on customers.
    + E.g. Name, address, phone number, etc.
    
## How is a database different than a collection of CSV files?

  - The acronym RDBMS stands for **Relational Database Management System**.
  - It consists of the database itself and the *software* necessary to its functionality.
  - In particular, a RDBMS has been optimized to perform certain tasks accurately and efficiently.
    + Data accuracy: make sure attributes satisfy certain conditions (e.g. a quantity should be non-negative).
    + Data retrieval: extract necessary data as quickly as possible.
    + Access: control who can change the data and who can query it.
    
## Databases and `R`

  - `R` allows us to connect to databases, retrieve information about the tables and attributes, and query the data.
  - You can treat the tables as `data.frame`s, use functions like `filter`, `mutate`, `summarise`, and `R` will transform your code into a query that the database can understand.
  - **Important principle**: RDBMS have been optimized to run queries fast, so we want to push as much of these computations to the database as we can, as opposed to extracting *all* the data into `R` and working directly with the `data.frame`s.
  
## Example {.allowframebreaks}

```{r}
library(DBI)
# duckdb is a type of RDBMS
library(duckdb)
```


```{r}
# 1. Create a connection
con <- dbConnect(duckdb())

# This database is actually empty...
# We fill it in with the data from nycflights13
library(nycflights13)
duckdb_register(conn = con, name = "flights", 
                df = flights)
```


```{r}
library(tidyverse)

tbl(con, "flights") %>% 
  group_by(dest) %>%
  summarise(delay = mean(dep_time, na.rm = TRUE))
```

```{r}
# To turn the output into a data frame
# use collect()
tbl(con, "flights") %>% 
  group_by(dest) %>%
  summarise(delay = mean(dep_time, na.rm = TRUE)) %>% 
  collect()
```

## Exercise {.allowframebreaks}

```{r}
# Let's revisit an example from last week
# where we looked at average delay

# First, add planes table
duckdb_register(conn = con, name = "planes", 
                df = planes)
```

```{r warning = FALSE}
# Replace data.frames by tbl(con, name)
data_avg <- left_join(tbl(con, "flights"),
                      tbl(con, "planes"),
                      by = "tailnum") %>% 
  mutate(tot_delay = dep_delay + arr_delay) %>% 
  group_by(year.y) %>% 
  summarise(avg_delay = mean(tot_delay)) %>% 
  collect() # Collect at the end for efficiency
```

```{r message = FALSE, warning = FALSE}
data_avg %>% 
  ggplot(aes(x = year.y,
             y = avg_delay)) +
  geom_point()
```

## Summary

\begin{enumerate}
  \item Create a connection to the database using \texttt{dbConnect}.
  \begin{itemize}
    \item Requires we know what type of RDBMS (e.g DuckDB, MySQL, Oracle) it is.
  \end{itemize}
  \item Work with tables like \texttt{data.frame}s using \texttt{tbl(con, name)}.
  \item Only once data manipulation is done, \texttt{collect} the results.
\end{enumerate}

## Live demo

<center>
For the live demo, I will use a SQLite database named `database.db` that you can find on UM Learn.
</center>
