---
title: "Regular expressions"
draft: true
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: SCI 2000--Introduction to Data Science
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(cache = FALSE, message = FALSE,
                      linewidth = 50)
```

## Lecture Objectives

  - Understand the definition of regular expressions
  - Recognize and use the different metacharacters
  - Use regular expressions to filter and edit text data

## Motivation

  - As we have seen, data is often encoded using text.
    + `M/F` or `minivan/suv/pickup`
  - When we are lucky, this data is free of errors and coded exactly the way we need it for data analysis.
  - More often than not, there are typos or inconsistencies in the data that need to be addressed.

## Regular expression--Definition

  - A **regular expression** (or **regex**) is a sequence of characters that specify a search pattern.
  - In other words, a regex is a pattern that we want to find in a string of text.
  - Common applications of regexes include:
    + finding observations where a certain word appears
    + replacing a string by another one
    + splitting a string according to a certain pattern

## Examples {.allowframebreaks}

```{r}
library(stringr)
# Detect a pattern
str_detect(c("apple", "orange", "pineapple"),
           pattern = "apple")
# Careful: This is case-sensitive
str_detect(c("Apple", "Orange", "Pineapple"),
           pattern = "apple")
# To ignore case
str_detect(c("Apple", "Orange", "Pineapple"),
           pattern = regex("apple", 
                           ignore_case = TRUE))
```

```{r}
# Match one of the patterns
str_detect(c("color", "colour", "coulour"),
           pattern = "color|colour")
# Or more compact
str_detect(c("color", "colour", "coulour"),
           pattern = "col(o|ou)r")
```

```{r}
# Replace patterns
str_replace_all(c("Male", "male", "MaLe"),
                pattern = regex("male", 
                                ignore_case = TRUE),
                replacement = "male")
```

\vspace{1cm}

```{r}
# Split a string using a pattern
str_split(c("it is a sentence", "it is another one"),
          pattern = " ")
```

## Anchors

  - **Anchors** are special characters (i.e. metacharacters) that can be used to specify *where* we want to find a match.
  - There are two main anchors:
    + `^pattern` will match any string that *starts* with `pattern`
    + `pattern$` will match any string that *ends* with `pattern`
  - You can combine them:
    + `^pattern$` will *only* match the string `pattern`
  - If you want to match on a metacharacter (e.g. `$`), you need to escape it (see example below).

## Example {.allowframebreaks}

```{r}
# This doesn't work...
str_detect(c("$15.99", "$3.75", "1.99$"), 
           pattern = "^$")
# But this does!
str_detect(c("$15.99", "$3.75", "1.99$"), 
           pattern = "^\\$")
```

```{r}
# Matching on prices that start or end with dollar sign
# Use logical operator
str_detect(c("$15.99", "$3.75", "1.99$"), 
           pattern = "^\\$") |
  str_detect(c("$15.99", "$3.75", "1.99$"), 
           pattern = "\\$$")
```

\vspace{2cm}

```{r}
# Or more compact
str_detect(c("$15.99", "$3.75", "1.99$"), 
           pattern = "^\\$|\\$$")
```

## Quantifiers

  - **Quantifiers** are ways to specify how many times a certain pattern should appear.
    + At least once? Exactly three times?
  - There are four important metacharacters to remember.
    + `.` will match any single character, except a new line.
    + `?` will match the item on its left at most once.
    + `*` will match the item on its left zero or more times.
    + `+` will match the item on its left once or more times.
  - Key distinction between `*` and `+`
    + the latter requires at least one match.
  
## Example {.allowframebreaks}

```{r}
# Revisiting an earlier example
str_detect(c("color", "colour", "coulour"),
           pattern = "colou?r")

# Matching strings that end with a bunch of periods
str_detect(c("str", "str.", "str..", "str..."),
           pattern = "\\.+$")
```

## Quantifiers cont'd

  - You can also control the number of matches more precisely.
    + `{n}` will match the item on its left exactly n times.
    + `{n,}` will match the item on its left at least n times.
    + `{n,m}` will match the item on its left at least n times, but no more than m times.

## Exercise

<center>
Find a regular expression that matches string ending with an ellipsis (i.e. three dots).
</center>

## Solution

```{r}
str_detect(c("string.", "string..", "string..."),
           pattern = "\\.{3}$")
# Be careful: a string with 4 dots will also match
str_detect("string....", pattern = "\\.{3}$")
```

## Character classes {.allowframebreaks}

  - When discussing quantifiers, I used "item on the left" instead of "character on the left".
  - This was intentional: these items could also be **character classes**.
  - We can create them using square brackets.
    + E.g. `p[ao]rt` will match both `part` and `port`.
  - Character classes can also be created using *sequences*.
    + `[a-z]` will match all lower case letters
    + `[a-zA-Z]` will match all lower and upper case letters
    + `[0-9]` will match all ten digits
    
\vspace{1cm}

  - There are also built-in character classes:
    + `\\d` matches any digit character (equivalent to `[0-9]`)
    + `\\s` matches any space character (including tabs, new lines, etc.)
    + `\\w` matches any word character (equivalent to `[A-Za-z0-9_]`)
    + `\\b` matches word boundaries
  - Finally, you can negate character classes to get non-matches.
    + `p[^ao]rt` matches `purt` and `pert`
    + The negation of `\\d`, `\\s`, `\\w`, `\\b` are `\\D`, `\\S`, `\\W`, `\\B` respectively.

## Example {.allowframebreaks}

```{r}
# Split a sentence into words
str_split("The fox ate a berry.", "\\b")
str_split("The fox ate a berry.", "\\s")
```

```{r}
# Trim white space
str_replace_all("Is  this     enough?", 
                pattern = "\\s+", 
                replacement = " ")
```

## Exercise

<center>
Find a regular expression that matches white space at the beginning and the end of a string. 
</center>

## Solution

```{r}
str_replace_all(" Is this enough?   ",
                pattern = "(^\\s+|\\s+$)",
                replacement = "")
```

## Summary

  - Regular expressions are patterns that we want to search within a string.
  - Anchors and quantifiers are metacharacters that allow us to be quite specific about the type of matches we want.
    + Remember: to match on a metacharacter literally, you need to escape it!
  - Most modern implementations of regular expressions also have **lookaround operators**, which we won't cover. But look it up if you're interested!
  
## Example {.allowframebreaks}

```{r}
library(tidyverse)
library(dslabs)
glimpse(movielens)
```

```{r}
# How many horror movie reviews
movielens %>% 
  filter(str_detect(genres, "Horror")) %>% 
  nrow()
```

----

```{r}
# What proportion are about thrillers?
movielens %>% 
  mutate(thriller = str_detect(genres,
                               "Thriller")) %>% 
  summarise(prop = mean(thriller))
```

----

```{r}
# What genre is Forrest Gump?
movielens %>% 
  filter(str_detect(title, "Gump")) %>% 
  pull(genres) %>% 
  unique %>%
  str_split(pattern = "\\|")
```

## Exercise

<center>
The dataset `reported_heights` in the `dslabs` package contains self-reported heights in no specific format. Clean up the data by making all heights comparable.

This is a challenging exercise. Try to do as much as you can.
</center>

## (Partial) solution {.allowframebreaks}

  - The first thing to do is to decide which units we will use.
  - Looking at the data, it seems that most people reported their height in inches.
    + Therefore, let's use inches as our unit of measurement.
  - Next, we need to look at the data and find heights that don't seem to be in inches.
  - For example, some heights are a single digit, which is probably the height in feet.
    + To find these heights, we can use `^\\d$` as our regex.
    + Once found, we need to convert to an integer and multiply by 12.
    
```{r}
library(tidyverse)
library(dslabs)
library(stringr)

reported_heights %>% 
  filter(str_detect(height, "^\\d$")) %>% 
  count(height)
```

  - We can see some errors already.
    + A height of 0 or 1 is probably a mistake.
    + A height of 2 is either a mistake or 2 meters. We will assume the latter.
    
```{r}
data_clean1 <- reported_heights %>% 
  filter(!height %in% c(0, 1)) %>% # Remove mistakes
  filter(str_detect(height, "^\\d$")) %>% 
  mutate(height_in = if_else(height == "2",
                             78.75,
                             12 * as.numeric(height)))
nrow(data_clean1)
```

  - We can look for other patterns by looking at heights that are not written as one or two digits.
    + The regex we want to use is `^\\d{1,2}$`
    
```{r}
reported_heights %>% 
    filter(!str_detect(height, "^\\d{1,2}$")) %>%
  head()
```

  - For a complete-ish solution, see UM Learn.