---
title: "Visualization Principles"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: SCI 2000--Introduction to Data Science
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(cache = FALSE, message = FALSE,
                      linewidth = 50)
```

## Lecture Objectives

  - Choose the right visual cues for your data visualization
  - Understand when the axis should include 0
  - Identify which visual cues can distort quantities
  - Emphasize important comparisons

## Motivation

  - In the previous lecture, we discussed different types of data visualizations.
  - We also discussed their pros and cons.
  - But we need general principles of effective data visualizations.
    + Better visualization = better communication

## Visual toolbox

  - Let's say we want to create a data visualization based on a single continuous variable
    + E.g. age, income, blood pressure. 
  - What is the best way to highlight *differences* between subgroups?
  - As data analysts, we have several visual cues we can use:
    + Position, angle, length, area, colour, angle
  - Not all visual cues are created equal! 
  - It turns out that humans are much better at understanding some visual cues (e.g. lengths) than others (e.g. volumes).
  
## Hierachy of visual cues

\begin{columns}
\begin{column}{0.5\textwidth}
   Researchers have found that there is a hierarchy of visual cues, from "accurate" to "generic", and we should aim to be as accurate as possible.
\end{column}
\begin{column}{0.5\textwidth}
    \begin{center}
     \includegraphics[width=\textwidth]{cue_hierarchy.jpg}
     \end{center}
\end{column}
\end{columns}

## Exercise

### What are the visual cues?

```{r, eval = TRUE, echo = FALSE}
library(tidyverse)
library(dslabs)
library(ggrepel)
library(cowplot)

west <- c("Western Europe", "Northern Europe", "Southern Europe",
          "Northern America", "Australia and New Zealand")

dat <- gapminder %>% 
    filter(year%in% c(2010, 2015) & region %in% west & 
               !is.na(life_expectancy) & population > 10^7) %>% 
    mutate(year = as.factor(year))

plot1 <- dat %>%
    ggplot(aes(year, life_expectancy, group = country)) +
    geom_line(aes(color = country), show.legend = FALSE) +
    geom_text_repel(data = ~filter(.x, year == 2015),
                    aes(label = country), x = 2,
                    nudge_x = 0.1, direction = "y", 
                    hjust = "left",
                    show.legend = FALSE) +
    xlab("") + ylab("Life Expectancy")

plot2 <- dat %>%
    dplyr::select(country, year, life_expectancy) %>% 
    spread(year, life_expectancy) %>% 
    mutate(diff = `2015` - `2010`) %>% 
    ggplot(aes(`2010`, `2015`)) +
    geom_point(aes(size = diff)) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed") +
    geom_text_repel(aes(label = country),
                    show.legend = FALSE) +
    xlab("2010") + ylab("2015") +
    scale_size(name = "Difference")

cowplot::plot_grid(plot2, plot1)
```

## Solution

  - Size and slope are used to highlight differences over time.
  - Position is used to highlight differences across countries.
  - Colour and text are used to identify countries.
  - Length/position (with respect to diagonal line) is also used to highlight differences.

## Pie charts {.allowframebreaks}

  - **Pie charts** form somewhat of a paradox: 
    + they are very popular; 
    + they are considered poor data visualizations. 
  - Let's look at an example:

```{r pie1, eval = TRUE, echo = FALSE, message = FALSE}
library(tidyverse)
# Browser preference
df <- data.frame(
  "Browser" = c("Opera", "Safari", "Firefox", "Chrome", "IE"),
  "Y2000" = c(3, 21, 23, 26, 28),
  "Y2015" = c(2, 22, 21, 29, 27)
)
# Can trick ggplot to make a pie chart:
# Make a bar chart in polar coordinates
ggplot(df, aes(x = "", y = Y2000, fill = Browser)) + 
  geom_bar(stat = "identity") + 
  coord_polar("y", start = 0) +
  theme_void()
```

  - Can you really tell which browser is the most popular?
  - Pie charts are difficult to read because *angles* can be hard to understand.
    + But they are also very hard to compare.
    
```{r pie3, eval = TRUE, echo = FALSE}
# gather combines Y2000 and Y2015 
# and adds a key column
df_tall <- gather(df, "Year", "Percent", 
                  Y2000, Y2015)
ggplot(df_tall, aes(x = "", y = Percent, fill = Browser)) + 
  geom_bar(stat = "identity") + 
  coord_polar("y", start = 0) +
  facet_grid(~ Year) +
  theme_void()
```

  - Can you tell if the preference ranking changed between 2000 and 2015? 
  - **Pie charts can almost always be replaced by a more useful chart** (or even a table).
  - In this case, a bar chart can more easily show us changes.
  
```{r pie4, eval = TRUE, echo = FALSE}
ggplot(df_tall, aes(x = Browser, y = Percent,
                    fill = Browser)) + 
  geom_bar(stat = "identity") + 
  facet_grid(~Year) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r pie5, eval = TRUE, echo = FALSE}
ggplot(df_tall, aes(x = Browser, y = Percent,
                    fill = Year)) + 
  geom_bar(stat = "identity",
           position = "dodge") +
  theme_minimal()
```

## Include Zero? {.allowframebreaks}

  - When constructing the barplots above, we made an important assumption: the length of the bar is proportional to the quantity being displayed. 
  - This assumption is important, because it was what allows to compare across browsers and years.         + Also, it allows us to assess the *relative* difference between them. 
  - When we don't include zero in a barplot, the lengths are no longer proportional to the quantities.
  
```{r zero1, eval = TRUE, echo = FALSE}
# Remove Opera from list of browsers
df_tall2 <- filter(df_tall, Browser != "Opera")
# Restrict the range
ggplot(df_tall2, aes(x = Browser, y = Percent,
                     fill = Browser)) + 
  geom_bar(stat = "identity") + 
  facet_grid(~Year) +
  coord_cartesian(ylim = c(20, 30)) +
  theme_minimal()
```

  - Differences are exaggerated when we remove the bottom of the bars. 
    + Looks like the popularity of Firefox has shrunk by 50%, but the difference is only 2%!
  - **When should we include zero?**
    + When the visual cue we use is **length** (e.g. barplots). 
  - When the visual cue is **position**, then excluding zero does not distort the quantity of interest.

```{r zero3, eval = TRUE, echo = FALSE}
library(dslabs)
gapminder %>% 
  filter(year == 2012) %>% # Only keep data from 2012
  ggplot(aes(continent, life_expectancy)) +
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(height = 0) +
  theme_minimal() +
  xlab("") + ylab("Life expectancy (in years)")
```

<!-- ## Distortion -->

<!-- The `gapminder` dataset also contains information about *fertility*. What if we wanted to add this information to the graph? One possibility would be to make the "size" of the points proportional to the value of fertility. -->

<!-- **But be careful**: Remember from geometry that the area of circle is proportional to the *square* of the radius! So if we make the *radius* proportional to fertility, a country with twice the fertility level of another country would be assigned a point that is *four times* the size of the other country. In other words, because the visual cue is the *area* (or size) of the point, we would be **distorting** the comparisons. -->

<!-- For this reason, `ggplot2` automatically makes the area of the points (and not the radius) proportional to the variable. Let's look at an example: -->

<!-- ```{r distort1, exercise = TRUE, exercise.eval = TRUE} -->
<!-- gapminder %>%  -->
<!--   filter(year == 2012) %>% # Only keep data from 2012 -->
<!--   ggplot(aes(continent, life_expectancy, -->
<!--              size = fertility)) + -->
<!--   geom_point() -->
<!-- ``` -->

<!-- The graph above does not distort the comparisons. For example, we can compare fertility across continents quite easily. -->

<!-- Here's what it would look like if we made the *radius* of the points proportional to fertility: -->

<!-- ```{r distort2, exercise = TRUE, exercise.eval = TRUE} -->
<!-- # THIS IS AN EXAMPLE OF WHAT NOT TO DO -->
<!-- gapminder %>%  -->
<!--   filter(year == 2012) %>%  -->
<!--   ggplot(aes(continent, life_expectancy, -->
<!--              size = fertility)) + -->
<!--   geom_point() + -->
<!--   scale_radius() # This tells ggplot2 to use radius instead of area -->
<!-- ``` -->

<!-- As we can see, comparisons are much harder: twice the fertility means a point that is four times bigger! -->

## Ordering categories {.allowframebreaks}

  - By default, `R` orders strings in alphabetical order. 
    + When is alphabetical order the best way to order information? 
    + Wouldn't you rather order Canadian provinces from West to East?
  - Reordering categories according to a more meaningful criterion can often improve your data visualization by making it easier to compare. 
  
```{r reorder1, eval = TRUE, echo = FALSE}
data_murders <- murders %>%
  mutate(murder_rate = total / population * 100000) # Compute rate

data_murders %>%
  ggplot(aes(state, murder_rate)) +
  geom_bar(stat = "identity") +
  coord_flip() + # For horizontal bars
  theme(axis.text.y = element_text(size = 6)) + # Shrink text
  xlab("")
```

  - We can easily identify the state with the highest rate (DC), but it's more difficult to identify the state with the smallest rate.
    + Is it HI, NH, or VT? 
  - More crucially, it is hard to identify states that have similar rates.
  - Instead, we can reorder the states according to the rate themselves. 
    + We can use the function `reorder` to arrange US states according to murder rate

```{r reorder3, eval = TRUE, echo = TRUE}
library(tidyverse)
library(dslabs)

data_murders %>%
  mutate(murder_rate = total / population * 100000) %>% # Compute rate
  mutate(state = reorder(state, murder_rate)) %>% 
  ggplot(aes(state, murder_rate)) +
  geom_bar(stat = "identity") +
  coord_flip() + # For horizontal bars
  theme(axis.text.y = element_text(size = 6)) +
  xlab("")
```

## Exercise

### Which is easier to interpret?

```{r datavizex1-setup, echo = FALSE, eval = TRUE}
gg1 <- us_contagious_diseases %>% 
  filter(disease == "Measles", year == 1928,
         !state %in% c("Hawaii", "Alaska")) %>% 
  mutate(rate = 10000*count/population) %>% 
  ggplot(aes(state, rate)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  xlab("")

gg2 <- us_contagious_diseases %>% 
  filter(disease == "Measles", year == 1928,
         !state %in% c("Hawaii", "Alaska")) %>% 
  mutate(rate = 10000*count/population,
         state = reorder(state, rate)) %>% 
  ggplot(aes(state, rate)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  xlab("")
gridExtra::grid.arrange(gg1, gg2, ncol = 2)
```


## Easier comparisons {.allowframebreaks}

  - How you arrange your visual cues can also make comparisons easier. 
  - Let's say we want to compare life expectancy between Americas and Europe.
    + E.g. We can use histograms.

```{r compare1, eval = TRUE, echo = TRUE}
library(tidyverse)
library(dslabs)

gapminder %>% 
  filter(year == 2012,
         continent %in% c("Americas", "Europe")) %>%
  ggplot(aes(life_expectancy,
             fill = continent)) +
  geom_histogram()
```

  - Can you determine which continent has the highest average? 
  - Let's plot two separate histograms.

```{r compare2, eval = TRUE, echo = TRUE}
gapminder %>% 
  filter(year == 2012,
         continent %in% c("Americas", "Europe")) %>%
  ggplot(aes(life_expectancy)) +
  geom_histogram() +
  facet_grid(~ continent)
```

  - Note: by default, `ggplot2` uses the same range for the axes to prevent distortions. 
  - But it is still difficult to determine which continent has the highest average... 
  - A better solution is to align the two histograms *vertically*.

```{r compare3, eval = TRUE, echo = TRUE}
gapminder %>% 
  filter(year == 2012,
         continent %in% c("Americas", "Europe")) %>%
  ggplot(aes(life_expectancy)) +
  geom_histogram() +
  facet_grid(continent ~ .) # Note: which side of ~ is important!
```

## Exercise 

<center>
Actually, it may even be easier to see the difference if we used boxplots. 

Create a plot with two side-by-side boxplots (one for Americas, one for Europe). 

Bonus: add the data points on top!
</center>

## Solution {.allowframebreaks}

```{r}
gapminder %>% 
  filter(year == 2012,
         continent %in% c("Americas", "Europe")) %>%
  ggplot(aes(continent, life_expectancy)) +
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(height = 0)
```


## Summary

  - Not all visual cues are created equal!
  - Pie charts are almost always a bad idea.
  - Include zero when length is an important visual cue.
  - Order your categories in a meaningful way.
  - When deciding how to arrange your visual cues on the graph, try to find the arrangement that makes the **important comparisons** as clear as possible.