---
title: "Data Visualization"
draft: true
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: SCI 2000--Introduction to Data Science
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(cache = FALSE, message = FALSE,
                      linewidth = 50)
```

## Lecture Objectives

  - Identify the main types of data visualization
  - Contrast their strengths and weaknesses

## Motivation

  - Summary statistics are useful in doing quick comparisons.
    + Or even statistical inference
  - Data visualizations are an effective way of sharing *a lot* of information about a dataset.
  - In this lecture, we'll focus on the main types of data visualizations; in the next lecture, we'll discuss important principles for **effective** visualization.

## Main principles

Why would we want to visualize data?

  - Quality control
  - Identify outliers
  - Find patterns of interest (EDA)
  - *Communicate results*
  
## Histogram {.allowframebreaks}

  - A **histogram** represents the frequency of observations occurring in certain bins.
    + Most software will choose default bins, but you can always change them.
  - It is useful for displaying continuous data, and comparing its distribution across subgroups.

```{r, message=FALSE}
library(tidyverse)
library(dslabs)

dim(olive)

# Create histogram for oleic acid
ggplot(olive,
       aes(x = oleic)) + 
  geom_histogram()
```

```{r hist_fill, message=FALSE}
# Look at distribution by region
ggplot(olive,
       aes(x = oleic, fill = region)) + 
  geom_histogram()
```


```{r hist_dodge, message=FALSE}
# Dodge instead of stack
ggplot(olive,
       aes(x = oleic, fill = region)) + 
  geom_histogram(position = "dodge")
```


```{r hist_facet, message=FALSE}
# Or with facets
ggplot(olive, 
       aes(x = oleic)) +
  geom_histogram() +
  facet_grid(. ~ region)
```

## Histogram--Summary

  - **Histograms** help visualize the distribution of a single variable.
    + It bins data and displays the counts in each bin
    + But large bins can hide important features, while small bins can create artifacts.
  - `ggplot` takes a `data.frame` as input and maps variables to different features of the graph.
    + `oleic` is mapped to the `x`-axis
    + `region` is mapped to the `fill` colour.
    + **Important**: This mapping happens inside the function `aes`.
  - `ggplot` automatically takes care of choosing the colour, drawing the limits, and printing a legend.
  - `facet_grid` can be used to display multiple plots together, one per value of the variable.

## A more complex histogram {.allowframebreaks}

```{r message = FALSE}
# Create a copy of the data to serve as background
olive_bg <- select(olive, -region)
ggplot(olive, aes(x = oleic)) + 
  # Start with grey background
  geom_histogram(data = olive_bg, 
                 fill = 'grey') +
  # Add colour on top
  geom_histogram(aes(fill = region)) +
  facet_grid(. ~ region) +
  # Move legend to top
  theme(legend.position = 'top')
```

## Exercise

<center>
Use the dataset `nba_players_19` from the package `openintro` to plot a histogram of the heights of basketball players.

Next, use histograms to compare the height distribution of guards vs centers.
</center>

## Solution {.allowframebreaks}

  - First, we plot the overall histogram.
  
```{r, message = FALSE}
library(tidyverse)
library(openintro)

ggplot(nba_players_19, aes(height)) +
  geom_histogram()
```

  - Next, we need to figure out which variable encodes the position of each player.
    + You can look at the help page `?nba_players_19`.
    + You can look at `str(nba_players_19)`.
  - Then we can filter using `position`.
  
```{r}
nba_players_19 %>% 
  filter(position %in% c("Center", "Guard")) %>% 
  ggplot(aes(height)) +
  geom_histogram() +
  facet_grid(~position)
```

## Density plot {.allowframebreaks}

  - **Density plots** can be thought of as *smoothed* histograms.
    + Their mathematical definition is much more involved and beyond the scope of this course.
  - They can be used interchangeably with histograms.

```{r, message=FALSE}
ggplot(olive, aes(x = oleic)) + 
  geom_density()

# Split by region
ggplot(olive, aes(x = oleic,
                  fill = region)) + 
  geom_density()

# Add transparency
ggplot(olive, aes(x = oleic,
                  fill = region)) + 
  geom_density(alpha = 0.5)

# Alternative: stacked density plots
ggplot(olive, aes(x = oleic,
                  fill = region)) + 
  geom_density(position = "stack")
```

## Density plot--Summary

  - **Density plots** can be thought of as *smoothed* histograms.
    + There is a parameter controlling the level of smoothness: too large and it will hide important features; too small and it may create artifacts.
  - We used a different *geom* to create the plot.
    + `geom_smooth` as opposed to `geom_histogram`.
  - The attribute `alpha` can be used to control transparency.
    + `alpha = 0` is completely transparent
    + `alpha = 1` is completely opaque.

## Boxplot {.allowframebreaks}

  - Box plots are a simple way to display important quantiles and identify outliers
  - Components (per Tukey):
    + A box delimiting the first and third quartile;
    + A line indicating the median;
    + Whiskers corresponding to the lowest datum still within 1.5 IQR of the lower quartile, and the highest datum still within 1.5 IQR of the upper quartile;
    + Any datum that falls outside the whiskers is considered a (potential) outlier.

```{r, message=FALSE}
ggplot(olive, aes(x = oleic)) + 
  geom_boxplot(y = 0) # y = 0 is a dummy value

# Map region to y-axis
ggplot(olive, aes(x = oleic, 
                  y = region)) + 
  geom_boxplot()

# Add all points on top of boxplots
ggplot(olive, aes(x = oleic, 
                  y = region)) + 
  geom_boxplot() +
  geom_point()

# Add vertical noise to the points to reduce overlap
# Note: need to remove outliers or you will get 
#       duplicates
ggplot(olive, aes(x = oleic, 
                  y = region)) + 
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(height = 0.25, width = 0) 

# Flip boxplots by switching the axes
ggplot(olive, aes(x = region, 
                  y = oleic)) + 
  geom_boxplot()
```

## Boxplot--Summary

  - **Boxplots** are a mixture between a data visualization and a summary statistics.
    + It is essentially a graphical depiction of the five-number summary.
  - Widely different datasets can give rise to the same boxplot.
    + I recommend to overlay the actual data.

## Exercise

<center>
Using the dataset `nba_players_19` from the package `openintro`, compare the distribution of heights across all positions.
</center>

## Solution {.allowframebreaks}

```{r}
ggplot(nba_players_19, aes(x = position,
                           y = height)) +
  geom_boxplot(outlier.colour = NA) +
  geom_jitter(height = 0, width = 0.25)
```


## Single-variable viz

  - All three data visualizations above focused on a single continuous variable.
  - But you can draw one such visualization for the same variable, but in different subgroups.
    + E.g. GPA for math, biology and psychology majors.
  - In this way, they can all be used to investigate the relationships between *one continuous and one categorical variable*.

# Bivariate plots

## Scatter plot {.allowframebreaks}

  - The simplest way to represent the relationship between two continuous variables is a **scatter plot**.
    + Not really suitable with categorical variables.
  - Technically still possible with three variables, but typically more difficult to read.
  
```{r}
ggplot(stars, aes(x = magnitude, 
                  y = temp)) + 
  geom_point()

# Add colour for type of stars
ggplot(stars, aes(x = magnitude, 
                  y = temp,
                  colour = type)) + 
  geom_point()
```

```{r, echo = FALSE, eval = FALSE}
library(scatterplot3d)

greenhouse_gases %>% 
  spread(gas, concentration) %>% 
  with(scatterplot3d(CH4,   # x axis
                     CO2,   # y axis
                     N2O    # z axis
))
```

## Exercise

<center>
Use the dataset `babies_crawl` from the package `openintro` to plot the average crawling age against the average outdoor temperature at 6 months.
</center>

## Solution {.allowframebreaks}

  - First, we need to figure out the name of the variables we need to plot.
    + You can look at the help page `?babies_crawl`.
    + You can look at `str(babies_crawl)`.
  - Our two variables are `temperature` and `avg_crawling_age`
  
```{r}
library(tidyverse)
library(openintro)

ggplot(babies_crawl, aes(x = temperature, 
                         y = avg_crawling_age)) +
  geom_point()
```

  - What if we want to restrict the range of temperatures?

```{r}
# First option
# Restrict the data before plotting
babies_crawl %>% 
    filter(temperature > 30, temperature < 70) %>% 
    ggplot(aes(x = temperature,
               y = avg_crawling_age)) +
    geom_point()
```


```{r linewidth = 50}
# Second option
# xlim removes the points from the plot
ggplot(babies_crawl, aes(x = temperature,
                         y = avg_crawling_age)) +
    geom_point() +
    xlim(c(30, 70))
```


```{r}
# Third option
# coord_cartesian zooms in/out 
ggplot(babies_crawl, aes(x = temperature,
                         y = avg_crawling_age)) +
    geom_point() +
    coord_cartesian(xlim = c(30, 70))
```

<!-- ## Bivariate density plot {.allowframebreaks} -->

```{r eval = FALSE}
ggplot(stars, aes(x = magnitude, 
                  y = temp)) + 
  geom_density_2d()

# We can add points on top of the contour lines
ggplot(stars, aes(x = magnitude, 
                  y = temp)) + 
  geom_density_2d() +
  geom_point()

# We can colour points by star type
# Note: colour is only defined for geom_point
ggplot(stars, aes(x = magnitude, 
                  y = temp)) + 
  geom_density_2d() +
  geom_point(aes(colour = type))
```

# Beyond two variables

## Limitations

  - Three-dimensional scatter plots are possible, but hard to interpret.
  - Density plots can technically be constructed for any dimension
    + But as the dimension increases, its performance *decreases* rapidly
  - **Solution**: We can look at each variable one at a time and at each pairwise comparison.
  
## Pairs plot {.allowframebreaks}

  - A pairs plot arranges these univariate summaries and pairwise comparisons along a matrix.
  - Each variable corresponds to both a row and a column
  - Univariate summaries appear on the diagonal, and pairwise comparisons off the diagonal.
  - Because of symmetry, we often see a different summary of the comparison above and below the diagonal.
  
```{r, message = FALSE}
library(GGally)

# Select three variables
olive_sub <- olive %>% 
  select(eicosenoic, arachidic, linolenic)

ggpairs(olive_sub)
```

\vspace{0.5in}

  - As we can see, `GGally` displays the following:
    + Scatter plots below the diagonal
    + Density plots on the diagonal
    + Pearson correlations above the diagonal
  - These can all be changed---see the documentation for more information.
